---
title: "dge"
author: "Jordan A. Lee"
date:  "2/28/2020 - `r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

This is a html version of dge.R as of end of day 2/28/20

# Background

DESeq2 doesn’t actually use normalized counts, rather it uses the raw counts and models the normalization inside the Generalized Linear Model (GLM). These normalized counts will be useful for downstream visualization of results, but cannot be used as input to DESeq2 or any other tools that peform differential expression analysis which use the negative binomial model.

`DESeq()` details:

+ Function runs: 1) est of size factors, est of dispersions, negative binomial generalized linear model (GLM) fitting and wald stats

`results()` details (these are outlined in the object printout below too): 

+ Adjusted pvals = BH corrected
+ User specified alpha val to use for adjusted pval thresholding for FDR analysis
+ Wald test to assess model fit (compared to likelihood ratio test). Wald test fits full model --> conducts tests for each gene to assess distance from 0 *- using the full model coefficent(s) and their standard error(s)* --> genes with coefficients significantly different than 0 will show this in their log2 fold change (and coefficient is reported as the log2fold change). - several studies have found that **Wald-log test** is better for RNA-seq data than others.


In RNA-seq data, larger variances grow with larger counts

# Analysis

```{r, cache=TRUE}
# Disclaimer: hardcoded names of patients in flat_matrix
# hard coded condition vector
####################
file_bulk1 <-'/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/01_process-bulkrna/07_ct_matrices/SCC_7319_B2_COUNTmatrix.txt'
file_bulk2 <-'/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/01_process-bulkrna/07_ct_matrices/SCC_7320_B2_COUNTmatrix.txt'
file_bulkGT1 <-'/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/01_process-bulkrna/07_ct_matrices/SCC_7319_GT_COUNTmatrix.txt'
file_bulkGT2 <-'/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/01_process-bulkrna/07_ct_matrices/SCC_7320_GT_COUNTmatrix.txt'
countsMECH_1_matrix_dir <- '/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/02_process-scrna/7319_Mech/'
countsENZ_1_matrix_dir <- '/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/02_process-scrna/7319_Enz/'
countsMECH_2_matrix_dir <- '/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/02_process-scrna/7320_Mech/'
countsENZ_2_matrix_dir <- '/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/02_process-scrna/7320_Enz/'
THRES = 0.1
output_sortedGenes = '/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/03_compare-bulk_scrna/scrna2bulk/geneOrdered_results.csv'

bulk_base_names <- c('bulk_7319', 'bulk_7320')
bulkgt_base_names <- c('bulkGT_7319', 'bulkGT_7320')
scmech_base_names <- c('sc_mech_7319', 'sc_mech_7320')
scenz_base_names <- c('sc_enz_7319', 'sc_enz_7320')
####################
```

```{r, message=FALSE}
library(Seurat)
library(dplyr)
library(Matrix)
library(gdata)
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(vsn)
library(RColorBrewer)
library(readr)
library(tibble)
library(tidyverse)
library(ggrepel)
```

# Create combined matrices of mech and enz 

Read in files and create pseudo-bulk (take the sum of raw gene counts across all cell barcodes and report that as the pseudo-bulk gene count). Preprocessing steps shown in `.Rmd`

```{r, cache=TRUE, echo=FALSE}
###############################
# Create combined count matrices
###############################
# Read in files
counts_mech1 <- Read10X(data.dir = countsMECH_1_matrix_dir, gene.column = 1)
counts_enz1 <- Read10X(data.dir = countsENZ_1_matrix_dir, gene.column = 1)
counts_mech2 <- Read10X(data.dir = countsMECH_2_matrix_dir,gene.column = 1)
counts_enz2 <- Read10X(data.dir = countsENZ_2_matrix_dir, gene.column = 1)
# convert to dense matrix
counts_mech1 <- as.matrix(counts_mech1) 
counts_enz1 <- as.matrix(counts_enz1) 
counts_mech2 <- as.matrix(counts_mech2) 
counts_enz2 <- as.matrix(counts_enz2) 
######## Subset for testing ###########
# counts_mech1 <- counts_mech1[256:260, 1:3]
# counts_mech1 <- counts_mech1[493:498, 1:3]
# counts_mech1
# counts_enz1 <- counts_enz1[493:498, 1:3]
# counts_enz1
# counts_mech2 <- counts_mech2[493:498, 1:3]
# counts_mech2
# counts_enz2 <- counts_enz2[493:498, 1:3]
# counts_enz2
#######################################
# Flaten counts (sum all cells for a gene) + combine matrices
mech1 <- as.data.frame(rowSums(counts_mech1))
enz1 <- as.data.frame(rowSums(counts_enz1)) 
mech2 <- as.data.frame(rowSums(counts_mech2)) 
enz2 <- as.data.frame(rowSums(counts_enz2)) 


########################
# Read in bulk samples and create a unified count matrix
########################
# Bulk
b1 <- read.table(file=file_bulk1, sep='\t', header=TRUE)
b2 <- read.table(file=file_bulk2, sep='\t', header=TRUE)
#combined bulk matrix
flat_matrix_BULK <- merge(b1, b2, by ='Geneid', all=TRUE ) #merge by rowname
colnames(flat_matrix_BULK) <- c('ensID',bulk_base_names)

# GT
gt1 <- read.table(file=file_bulkGT1, sep='\t', header=TRUE)
gt2 <- read.table(file=file_bulkGT2, sep='\t', header=TRUE)
#combined bulk matrix
flat_matrix_BULK_GT <- merge(gt1, gt2, by ='Geneid', all=TRUE ) #merge by rowname
colnames(flat_matrix_BULK_GT) <- c('ensID',bulkgt_base_names)
```



# Part 1. Establishing time and temp impact (bulk RNA-seq vs ground truth bulk RNA-seq)

```{r, message=FALSE, echo=FALSE}
############################
# Create count matrices for DGE
############################
# Combine matrices
flat_matrix_ctrlBULK <- merge(flat_matrix_BULK, flat_matrix_BULK_GT, by='ensID', all=TRUE)
# Replace NA --> 0 
flat_matrix_ctrlBULK[is.na(flat_matrix_ctrlBULK)]=0
# Now that have combined properly fix so that rowname are genes
flat_matrix_ctrlBULK <- column_to_rownames(flat_matrix_ctrlBULK, var='ensID')

############################
# Create meta data table of which samples are which treatment - needed for Deseq2
#############################
# Grab samples in order
samples <- colnames(flat_matrix_ctrlBULK)
# Create condition vector and type
condition <- c('bulk','bulk', 'gt', 'gt')
meta_data <- as.data.frame(condition)
rownames(meta_data) <- samples

# # Remove gene rows that are lowly expr in samples - filter low counts <10 ct
# flat_matrix_ctrlBULK <- flat_matrix_ctrlBULK[rowSums(flat_matrix_ctrlBULK > 10) != 0, ]

#####################
# DGE
#####################
dds_obj_ctrlBULK <- DESeqDataSetFromMatrix(countData = flat_matrix_ctrlBULK, colData = meta_data, design = ~ condition)
# DGE based on neg binom distr
dds_ctrlBULK <- DESeq(dds_obj_ctrlBULK, test ='Wald')
res_ctrlBULK <- results(dds_ctrlBULK, alpha=THRES, pAdjustMethod = 'BH', test='Wald')
( mcols(res_ctrlBULK)$description ) # info

###############
# Examine results
###############
#MA plot
print('MA plot of non-transformed counts')
plotMA(res_ctrlBULK, ylim=c(-5,5), alpha = THRES)
```

We can see that there are no signficantlly DE genes (would be denoted by red point).





# Part 2. Comparing scRNA-mech pseudo, scRNA-enz pseudo, and bulk - normalized

DGE analysis based on negative binomial distribution, mean and variance linked by local regression, can be implemented to estimate scale factors too. **Normalization method = DESeq2 median of ratios** which is the counts divided by sample size factors determined by meidan ratio of gene conts relative to geometric mean per gene - which accounts for seq depth and RNA composition. But not **gene length**

```{r, message=FALSE, echo=FALSE}
############################
# Create count matrices for DGE
############################
# Combine mech samples
flat_matrix_mech <- merge(mech1, mech2, by =0, all=TRUE) #merge by rowname
colnames(flat_matrix_mech) <- c("ensID",scmech_base_names)
# Combine enz samples
flat_matrix_enz <- merge(enz1, enz2, by =0, all=TRUE) #merge by rowname
colnames(flat_matrix_enz) <- c("ensID",scenz_base_names)
# Combine mech and enz samples
flat_matrix_ALL <- merge(flat_matrix_mech, flat_matrix_enz, by ='ensID', all=TRUE) #merge by rowname
# Combine mech, enz, bulk
flat_matrix_ALL <- merge(flat_matrix_BULK, flat_matrix_ALL, by='ensID', all=TRUE)
# Replace NA --> 0 
flat_matrix_ALL[is.na(flat_matrix_ALL)]=0
# Now that have combined properly fix so that rowname are genes
flat_matrix_ALL <- column_to_rownames(flat_matrix_ALL, var='ensID')


############################
# Create meta data table of which samples are which treatment - needed for Deseq2
#############################
# Grab samples in order
samples <- colnames(flat_matrix_ALL)
# Create condition vector and type
condition <- c('bulk','bulk', 'mech', 'mech', 'enz', 'enz')
meta_data <- as.data.frame(condition)
rownames(meta_data) <- samples

#### consider removing
# # Remove gene rows that are lowly expr in samples - filter low counts <10 ct
# flat_matrix_ctrlBULK <- flat_matrix_ctrlBULK[rowSums(flat_matrix_ctrlBULK > 10) != 0, ]
```

```{r, message=FALSE, echo=FALSE}
#####################
# DGE
#####################

dds_obj_ALL <- DESeqDataSetFromMatrix(countData = flat_matrix_ALL, colData = meta_data, design = ~ condition)
# DGE based on neg binom distr
dds_ALL <- DESeq(dds_obj_ALL, test ='Wald')
res_ALL <- results(dds_ALL, alpha=THRES, pAdjustMethod = 'BH', test='Wald')
( mcols(res_ALL)$description ) # info

#MA plot
print('MA plot of normalized counts')
plotMA(res_ALL, ylim=c(-5,5), alpha = THRES)

# "Table" of top diff expr
print('Most differ expr genes of normalized counts')
topgenes <- rownames(res_ALL)[res_ALL$padj <= sort(res_ALL$padj)[25] &!is.na(res_ALL$padj)]
print(topgenes)
```

Now let's look at the log10(counts) of the data

```{r, echo=FALSE}
##############
# Visualization
##############
# 1A. Extract normalized count matrix used by DESeq
    # estimating size factors is auto in DESEQ() but 
    # we have to do this manually to extract the normalized count matrix it produces
    # Normalization method= deseq median or ratios method
dds <- DESeqDataSetFromMatrix(countData = flat_matrix_ALL, colData = meta_data, design = ~ condition)
dds<- estimateSizeFactors(dds)
('size factors applied for norm were:')
(sizeFactors(dds)) #good for record keeping
normalized_counts_tibble <- counts(dds, normalized=TRUE)
#write.table(normalized_counts_tibble,file="/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/03_compare-bulk_scrna/scrna2bulk/normalized_counts--ALL.txt", sep="\t", quote=F, col.names=NA)

# 1B. get subset of top genes to look at, extract this info from DESeq then subset normalized count matrix
# Reformat
m_tibble <- meta_data %>% 
  rownames_to_column(var="samplename") %>% 
  as_tibble()
normalized_counts <- normalized_counts_tibble %>% 
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()
# Extract
top25_norm <- normalized_counts %>%
        filter(gene %in% topgenes)


# 2. Plot signficant DE genes
# Gathering the columns to have normalized counts to a single column (aka create a col sample name)
gathered <- top25_norm %>%
  gather(colnames(top25_norm)[2:7], key = "samplename", value = "normalized_counts") #2:7 bc headers are gene,then all 6 samples
# melt so that plot will be colored by condition
gathered <- inner_join(m_tibble, gathered,by="samplename")
ggplot(gathered) +
        geom_point(aes(x = gene, y = normalized_counts, color = condition)) +
        scale_y_log10() +
        xlab("Genes (Ranked)") +
        ylab("log10 Normalized Counts") +
        ggtitle("Top 25 Significant DE Genes") +
        theme_bw() +
	theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
	theme(plot.title = element_text(hjust = 0.5))

# 3. Heatmap
#plot only signficant genes and display their normalized expression
print('Heatmap of all normalized counts')
# Reformat 
annotation <- m_tibble %>%data.frame(row.names = "samplename")
mat<- top25_norm %>% column_to_rownames(var = "gene")
pheatmap(log10(mat), 
         cluster_rows = T,
         cluster_cols = T,
         show_rownames = T,
         border_color = NA,
         annotation = annotation, 
         fontsize = 5)
```


# Part 3. Comparing scRNA-mech pseudo vs bulk - normalized

```{r, message=FALSE, echo=FALSE}
############################
# Create count matrices for DGE
############################
flat_matrix_mech <- merge(mech1, mech2, by =0, all=TRUE) #merge by rowname
colnames(flat_matrix_mech) <- c("ensID",scmech_base_names)
# Add bulk samples in
flat_matrix_mech <- merge(flat_matrix_BULK, flat_matrix_mech, by='ensID', all=TRUE)
# Replace NA --> 0 
flat_matrix_mech[is.na(flat_matrix_mech)]=0
# Now that have combined properly fix so that rowname are genes
flat_matrix_mech <- column_to_rownames(flat_matrix_mech, var='ensID')

############################
# Create meta data table of which samples are which treatment - needed for Deseq2
#############################
# Grab samples in order
samples <- colnames(flat_matrix_mech)
# Create condition vector and type
condition <- c('bulk','bulk', 'mech', 'mech')
meta_data <- as.data.frame(condition)
rownames(meta_data) <- samples

# # Remove gene rows that are lowly expr in samples - filter low counts <10 ct
# flat_matrix_mech <- flat_matrix_mech[rowSums(flat_matrix_mech > 10) != 0, ]

#####################
# DGE
#####################
dds_obj_mech <- DESeqDataSetFromMatrix(countData = flat_matrix_mech, colData = meta_data, design = ~ condition)
# DGE based on neg binom distr
dds_mech <- DESeq(dds_obj_mech, test ='Wald')
res_mech <- results(dds_mech, alpha=THRES, pAdjustMethod = 'BH', test='Wald')
( mcols(res_mech)$description ) # info

#MA plot
print('MA plot of normalized counts')
plotMA(res_mech, ylim=c(-5,5), alpha = THRES)

# "Table" of top diff expr
print('Most differ expr genes of normalized counts')
topgenes <- rownames(res_mech)[res_mech$padj <= sort(res_mech$padj)[25] &!is.na(res_mech$padj)]
print(topgenes)
```

Now let's look at the log10(counts) of the data

```{r , echo=FALSE}
##############
# Visualization
##############
# 1A. Extract normalized count matrix used by DESeq
    # estimating size factors is auto in DESEQ() but 
    # we have to do this manually to extract the normalized count matrix it produces
    # Normalization method= deseq median or ratios method
dds <- DESeqDataSetFromMatrix(countData = flat_matrix_mech, colData = meta_data, design = ~ condition)
dds<- estimateSizeFactors(dds)
('size factors applied for norm were:')
(sizeFactors(dds)) #good for record keeping
normalized_counts_tibble <- counts(dds, normalized=TRUE)
#write.table(normalized_counts_tibble,file="/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/03_compare-bulk_scrna/scrna2bulk/normalized_counts--ALL.txt", sep="\t", quote=F, col.names=NA)
# 1B. get subset of top genes to look at, extract this info from DESeq then subset normalized count matrix
# Reformat
m_tibble <- meta_data %>% 
  rownames_to_column(var="samplename") %>% 
  as_tibble()
normalized_counts <- normalized_counts_tibble %>% 
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()
# Extract
top25_norm <- normalized_counts %>%
        filter(gene %in% topgenes)

# 2. Plot signficant DE genes
# Gathering the columns to have normalized counts to a single column (aka create a col sample name)
gathered <- top25_norm %>%
  gather(colnames(top25_norm)[2:5], key = "samplename", value = "normalized_counts") #2:7 bc headers are gene,then all 6 samples
# melt so that plot will be colored by condition
gathered <- inner_join(m_tibble, gathered,by="samplename")
ggplot(gathered) +
        geom_point(aes(x = gene, y = normalized_counts, color = condition)) +
        scale_y_log10() +
        xlab("Genes (Ranked)") +
        ylab("log10 Normalized Counts") +
        ggtitle("Top 25 Significant DE Genes") +
        theme_bw() +
	theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
	theme(plot.title = element_text(hjust = 0.5))
# # 3. Heatmap - SKIPPING FOR NOW BC COVARIANCE=0 IN SOME CASES - SEE USING cor(log(mat))
# #plot only signficant genes and display their normalized expression
# print('Heatmap of all normalized counts')
# # Reformat 
# annotation <- m_tibble %>%data.frame(row.names = "samplename")
# mat<- top25_norm %>% column_to_rownames(var = "gene")
# pheatmap(log10(mat), 
#          cluster_rows = T,
#          cluster_cols = T,
#          show_rownames = T,
#          border_color = NA,
#          annotation = annotation, 
#          fontsize = 5)
```


# Part 4. Comparing scRNA-enz pseudo vs bulk

```{r, message=FALSE, echo=FALSE}
############################
# Create count matrices for DGE
############################
flat_matrix_enz <- merge(enz1, enz2, by =0, all=TRUE) #merge by rowname
colnames(flat_matrix_enz) <- c("ensID",scenz_base_names)
# Add bulk samples in
flat_matrix_enz <- merge(flat_matrix_BULK, flat_matrix_enz, by='ensID', all=TRUE)
# Replace NA --> 0 
flat_matrix_enz[is.na(flat_matrix_enz)]=0
# Now that have combined properly fix so that rowname are genes
flat_matrix_enz <- column_to_rownames(flat_matrix_enz, var='ensID')

############################
# Create meta data table of which samples are which treatment - needed for Deseq2
#############################
# Grab samples in order
samples <- colnames(flat_matrix_enz)
# Create condition vector and type
condition <- c('bulk','bulk', 'enz', 'enz')
meta_data <- as.data.frame(condition)
rownames(meta_data) <- samples

# # Remove gene rows that are lowly expr in samples - filter low counts <10 ct
# flat_matrix_enz <- flat_matrix_enz[rowSums(flat_matrix_enz > 10) != 0, ]

#####################
# DGE
#####################
dds_obj_enz <- DESeqDataSetFromMatrix(countData = flat_matrix_enz, colData = meta_data, design = ~ condition)
# DGE based on neg binom distr
dds_enz <- DESeq(dds_obj_enz, test ='Wald')
res_enz <- results(dds_enz, alpha=THRES, pAdjustMethod = 'BH', test='Wald')
( mcols(res_enz)$description ) # info

#MA plot
print('MA plot of normalized counts')
plotMA(res_enz, ylim=c(-5,5), alpha = THRES)

# "Table" of top diff expr
print('Most differ expr genes of normalized counts')
topgenes <- rownames(res_enz)[res_enz$padj <= sort(res_enz$padj)[25] &!is.na(res_enz$padj)]
print(topgenes)
```

Now let's look at the log10(counts) of the data

```{r, echo=FALSE}
##############
# Visualization
##############
# 1A. Extract normalized count matrix used by DESeq
    # estimating size factors is auto in DESEQ() but 
    # we have to do this manually to extract the normalized count matrix it produces
    # Normalization method= deseq median or ratios method
dds <- DESeqDataSetFromMatrix(countData = flat_matrix_enz, colData = meta_data, design = ~ condition)
dds<- estimateSizeFactors(dds)
('size factors applied for norm were:')
(sizeFactors(dds)) #good for record keeping
normalized_counts_tibble <- counts(dds, normalized=TRUE)
#write.table(normalized_counts_tibble,file="/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/03_compare-bulk_scrna/scrna2bulk/normalized_counts--ALL.txt", sep="\t", quote=F, col.names=NA)
# 1B. get subset of top genes to look at, extract this info from DESeq then subset normalized count matrix
# Reformat
m_tibble <- meta_data %>% 
  rownames_to_column(var="samplename") %>% 
  as_tibble()
normalized_counts <- normalized_counts_tibble %>% 
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()
# Extract
top25_norm <- normalized_counts %>%
        filter(gene %in% topgenes)

# 2. Plot signficant DE genes
# Gathering the columns to have normalized counts to a single column (aka create a col sample name)
gathered <- top25_norm %>%
  gather(colnames(top25_norm)[2:5], key = "samplename", value = "normalized_counts") #2:7 bc headers are gene,then all 6 samples
# melt so that plot will be colored by condition
gathered <- inner_join(m_tibble, gathered,by="samplename")
ggplot(gathered) +
        geom_point(aes(x = gene, y = normalized_counts, color = condition)) +
        scale_y_log10() +
        xlab("Genes (Ranked)") +
        ylab("log10 Normalized Counts") +
        ggtitle("Top 25 Significant DE Genes") +
        theme_bw() +
	theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
	theme(plot.title = element_text(hjust = 0.5))
# # 3. Heatmap - SKIPPING FOR NOW BC COVARIANCE=0 IN SOME CASES - SEE USING cor(log(mat))
# #plot only signficant genes and display their normalized expression
# print('Heatmap of all normalized counts')
# # Reformat 
# annotation <- m_tibble %>%data.frame(row.names = "samplename")
# mat<- top25_norm %>% column_to_rownames(var = "gene")
# pheatmap(log10(mat), 
#          cluster_rows = T,
#          cluster_cols = T,
#          show_rownames = T,
#          border_color = NA,
#          annotation = annotation, 
#          fontsize = 5)
```
