---
title: "dge"
author: "Jordan Lee"
date: "2/28/2020"
output: html_document
---

This is a html version of dge.R as of end of day 2/28/20

# Background

`DESeq()` details:

+ Function runs: 1) est of size factors, est of dispersions, negative binomial generalized linear model (GLM) fitting and wald stats

`results()` details (these are outlined in the object printout below too): 

+ Adjusted pvals = BH corrected
+ User specified alpha val to use for adjusted pval thresholding for FDR analysis
+ Wald test to assess model fit (compared to likelihood ratio test). Wald test fits full model --> conducts tests for each gene to assess distance from 0 *- using the full model coefficent(s) and their standard error(s)* --> genes with coefficients significantly different than 0 will show this in their log2 fold change (and coefficient is reported as the log2fold change). - several studies have found that **Wald-log test** is better for RNA-seq data than others.

# Analysis

```{r, cache=TRUE}
# Disclaimer: hardcoded names of patients in flat_matrix
# hard coded condition vector
####################
file_bulk1 <-'/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/01_process-bulkrna/07_ct_matrices/SCC_7319_B2_COUNTmatrix.txt'
file_bulk2 <-'/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/01_process-bulkrna/07_ct_matrices/SCC_7320_B2_COUNTmatrix.txt'
countsMECH_1_matrix_dir <- '/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/02_process-scrna/7319_Mech/'
countsENZ_1_matrix_dir <- '/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/02_process-scrna/7319_Enz/'
countsMECH_2_matrix_dir <- '/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/02_process-scrna/7320_Mech/'
countsENZ_2_matrix_dir <- '/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/02_process-scrna/7320_Enz/'
THRES = 0.1
output_sortedGenes = '/Users/leejor/Ellrott_Lab/04_CSBC/cell-dissociation/data/03_compare-bulk_scrna/scrna2bulk/geneOrdered_results.csv'

bulk_base_names <- c('bulk_7319', 'bulk_7320')
scmech_base_names <- c('sc_mech_7319', 'sc_mech_7320')
scenz_base_names <- c('sc_enz_7319', 'sc_enz_7320')
####################
```

```{r, message=FALSE}
library(Seurat)
library(dplyr)
library(Matrix)
library(gdata)
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(vsn)
library(RColorBrewer)
library(readr)
library(tibble)
```

# Create combined matrices of mech and enz 

Read in files and create pseudo-bulk (take the sum of raw gene counts across all cell barcodes and report that as the pseudo-bulk gene count). Preprocessing steps shown in `.Rmd`

```{r, cache=TRUE, echo=FALSE}
###############################
# Create combined count matrices
###############################
# Read in files
counts_mech1 <- Read10X(data.dir = countsMECH_1_matrix_dir, gene.column = 1)
counts_enz1 <- Read10X(data.dir = countsENZ_1_matrix_dir, gene.column = 1)
counts_mech2 <- Read10X(data.dir = countsMECH_2_matrix_dir,gene.column = 1)
counts_enz2 <- Read10X(data.dir = countsENZ_2_matrix_dir, gene.column = 1)
# convert to dense matrix
counts_mech1 <- as.matrix(counts_mech1) 
counts_enz1 <- as.matrix(counts_enz1) 
counts_mech2 <- as.matrix(counts_mech2) 
counts_enz2 <- as.matrix(counts_enz2) 
######## Subset for testing ###########
# counts_mech1 <- counts_mech1[256:260, 1:3]
# counts_mech1 <- counts_mech1[493:498, 1:3]
# counts_mech1
# counts_enz1 <- counts_enz1[493:498, 1:3]
# counts_enz1
# counts_mech2 <- counts_mech2[493:498, 1:3]
# counts_mech2
# counts_enz2 <- counts_enz2[493:498, 1:3]
# counts_enz2
#######################################
# Flaten counts (sum all cells for a gene) + combine matrices
mech1 <- as.data.frame(rowSums(counts_mech1))
enz1 <- as.data.frame(rowSums(counts_enz1)) 
mech2 <- as.data.frame(rowSums(counts_mech2)) 
enz2 <- as.data.frame(rowSums(counts_enz2)) 


########################
# Read in bulk samples and create a unified count matrix
########################
b1 <- read.table(file=file_bulk1, sep='\t', header=TRUE)
b2 <- read.table(file=file_bulk2, sep='\t', header=TRUE)
# Combined bulk matrix
flat_matrix_BULK <- merge(b1, b2, by ='Geneid', all=TRUE ) #merge by rowname
colnames(flat_matrix_BULK) <- c('ensID',bulk_base_names)
```


# Part 1. Comparing scRNA-mech pseudo vs bulk

```{r, message=FALSE, echo=FALSE}
############################
# Create count matrices for DGE
############################
flat_matrix_mech <- merge(mech1, mech2, by =0, all=TRUE) #merge by rowname
colnames(flat_matrix_mech) <- c("ensID",scmech_base_names)
# Add bulk samples in
flat_matrix_mech <- merge(flat_matrix_BULK, flat_matrix_mech, by='ensID', all=TRUE)
# Replace NA --> 0 
flat_matrix_mech[is.na(flat_matrix_mech)]=0
# Now that have combined properly fix so that rowname are genes
flat_matrix_mech <- column_to_rownames(flat_matrix_mech, var='ensID')

############################
# Create meta data table of which samples are which treatment - needed for Deseq2
#############################
# Grab samples in order
samples <- colnames(flat_matrix_mech)
# Create condition vector and type
condition <- c('bulk','bulk', 'mech', 'mech')
meta_data <- as.data.frame(condition)
rownames(meta_data) <- samples

#####################
# DGE
#####################
# Remove gene rows that are lowly expr in samples - filter low counts <10 ct
flat_matrix_mech <- flat_matrix_mech[rowSums(flat_matrix_mech > 10) != 0, ]
dds_obj_mech <- DESeqDataSetFromMatrix(countData = flat_matrix_mech, colData = meta_data, design = ~ condition)
# DGE based on neg binom distr
dds_mech <- DESeq(dds_obj_mech, test ='Wald')
res_mech <- results(dds_mech, alpha=THRES, pAdjustMethod = 'BH', test='Wald')
( mcols(res_mech)$description ) # info

###############
# Examine results
###############
#MA plot
print('MA plot of non-transformed counts')
plotMA(res_mech, ylim=c(-5,5), alpha = THRES)
# "Table" of top diff expr
print('Most differ expr genes of non-transformed counts')
topgenes <- rownames(res_mech)[res_mech$padj <= sort(res_mech$padj)[25] &!is.na(res_mech$padj)]
print(topgenes)
# Heatmap - use transformed counts
print('Heatmap of all transformed counts')
rld <- rlogTransformation(dds_mech, blind=FALSE)
sig.dat <- assay(rld)[res_mech$padj < THRES & !is.na(res_mech$padj), ]
annC <- data.frame(condition=condition)
rownames(annC) <- colnames(sig.dat)
pheatmap(sig.dat, scale='row', fontsize_row=9,annotation_col = annC, show_rownames = FALSE, cluster_cols = TRUE, cluster_rows=TRUE)
```


# Part 2. Comparing scRNA-enz pseudo vs bulk

```{r, message=FALSE, echo=FALSE}
############################
# Create count matrices for DGE
############################
flat_matrix_enz <- merge(enz1, enz2, by =0, all=TRUE) #merge by rowname
colnames(flat_matrix_enz) <- c("ensID",scenz_base_names)
# Add bulk samples in
flat_matrix_enz <- merge(flat_matrix_BULK, flat_matrix_enz, by='ensID', all=TRUE)
# Replace NA --> 0 
flat_matrix_enz[is.na(flat_matrix_enz)]=0
# Now that have combined properly fix so that rowname are genes
flat_matrix_enz <- column_to_rownames(flat_matrix_enz, var='ensID')

############################
# Create meta data table of which samples are which treatment - needed for Deseq2
#############################
# Grab samples in order
samples <- colnames(flat_matrix_enz)
# Create condition vector and type
condition <- c('bulk','bulk', 'enz', 'enz')
meta_data <- as.data.frame(condition)
rownames(meta_data) <- samples

#####################
# DGE
#####################
# Remove gene rows that are lowly expr in samples - filter low counts <10 ct
flat_matrix_enz <- flat_matrix_eenz[rowSums(flat_matrix_enz > 10) != 0, ]
dds_obj_enz <- DESeqDataSetFromMatrix(countData = flat_matrix_enz, colData = meta_data, design = ~ condition)
# DGE based on neg binom distr
dds_enz <- DESeq(dds_obj_enz, test ='Wald')
res_enz <- results(dds_enz, alpha=THRES, pAdjustMethod = 'BH', test='Wald')
( mcols(res_enz)$description ) # info

###############
# Examine results
###############
#MA plot
print('MA plot of non-transformed counts')
plotMA(res_enz, ylim=c(-5,5), alpha = THRES)
# "Table" of top diff expr
print('Most differ expr genes of non-transformed counts')
topgenes <- rownames(res_enz)[res_enz$padj <= sort(res_enz$padj)[25] &!is.na(res_enz$padj)]
print(topgenes)
# Heatmap - use transformed counts
print('Heatmap of all transformed counts')
rld <- rlogTransformation(dds_enz, blind=FALSE)
sig.dat <- assay(rld)[res_enz$padj < THRES & !is.na(res_enz$padj), ]
annC <- data.frame(condition=condition)
rownames(annC) <- colnames(sig.dat)
pheatmap(sig.dat, scale='row', fontsize_row=9,annotation_col = annC, show_rownames = FALSE, cluster_cols = TRUE, cluster_rows=TRUE)
```

